---
title: "Spatial Analysis of Madrid Traffic Data with MetricGraph"
author: "David Bolin and Alexandre Simas - INLA Workshop 2025"
date: "May 2025"
output:
  pdf_document:
    number_sections: true
    fig_width: 8
    fig_height: 6
    highlight: tango
  html_document:
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "80%")
```

# Introduction

This tutorial explores spatial analysis of traffic data on road networks using the `MetricGraph` package. We'll work with real traffic data from Madrid, Spain, and learn how to:

1. Construct a metric graph from OpenStreetMap road network data
2. Add observations to the graph
3. Visualize traffic data on the network
4. Fit spatial models to analyze traffic patterns

The `MetricGraph` package provides tools for statistical analysis of data on network-structured domains, like road networks. It implements various types of Gaussian random fields on graphs and offers interfaces to INLA for Bayesian inference.

Let's begin by loading the necessary packages:

```{r load-packages}
library(data.table)
library(sf)
library(osmdata)
library(dplyr)
library(MetricGraph)
library(ggplot2)
library(inlabru)
```

# Data Description

For this tutorial, we've prepared datasets containing traffic information for Madrid. The datasets include:

1. Traffic sensor locations in Madrid
2. Traffic measurements from these sensors during evening rush hour (6-7 PM) in March 2025

These data have been generated based on typical traffic patterns and distributions for educational purposes.

## Loading the Data

For convenience, we've prepared data files containing:
- The traffic sensor locations
- The traffic measurements
```

Now we can load the pre-prepared data:

```{r load-data}
# Load the pre-prepared data files
madrid_traffic_avg_sf <- readRDS("madrid_traffic_avg.rds")
radar_locations_sf <- readRDS("radar_locations.rds")

# Examine the data
head(madrid_traffic_avg_sf)
```

# Madrid Road Network as a Metric Graph

## Understanding the Study Area

For this tutorial, we're focusing on a central area of Madrid defined by this bounding box:

```{r define-bbox}
# Define the Madrid bounding box used for this tutorial
small_bbox <- c(-3.73, 40.40, -3.65, 40.45)  

# Create a simple data frame for the bounding box to visualize it
bbox_df <- data.frame(
  lon = c(small_bbox[1], small_bbox[3], small_bbox[3], small_bbox[1], small_bbox[1]),
  lat = c(small_bbox[2], small_bbox[2], small_bbox[4], small_bbox[4], small_bbox[2])
)
bbox_sf <- st_as_sf(bbox_df, coords = c("lon", "lat"), crs = 4326) %>%
  st_combine() %>%
  st_cast("POLYGON")
```


## Exercise 1: Exploring the Graph Structure

Let us start by creating a graph from the OpenStreetMap data with some standard features.

```{r create-graph-example, eval=FALSE}
madrid_sf_filtered <- opq(bbox = small_bbox) %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary")) %>%
  osmdata_sf()

madrid_graph_components <- graph_components$new(madrid_sf_filtered, perform_merges = TRUE, 
tolerance = list(vertex_vertex = 0.001,
vertex_edge = 0.01))

madrid_graph <- madrid_graph_components$get_largest()
```

Let's explore the pre-loaded graph structure:

```{r explore-graph}
# Look at the graph structure
print(madrid_graph)

# What is the total length of the road network in the graph?
cat("Total length of road network:", round(sum(madrid_graph$edge_lengths), 2), "km\n")

# Compute additional characteristics
madrid_graph$compute_characteristics()
print(t(madrid_graph$characteristics))
```

## Visualizing the Graph

Let's visualize our road network graph:

```{r plot-graph}
# Plot the graph
p <- madrid_graph$plot(vertex_size = 0.5, edge_width = 0.5)
print(p)
```

# Traffic Data on the Road Network

## Examining Sensor Locations 

Let's examine the traffic sensor locations in our study area:

```{r examine-sensors}
# How many sensors are in our study area?
cat("Number of sensors in the study area:", nrow(radar_locations_sf), "\n")

# Plot the graph with the sensor locations
p <- madrid_graph$plot(vertex_size = 0, edge_width = 0.5)
p + geom_sf(data = radar_locations_sf, color = "red", size = 1)
```

## Recreate the graph with more features

Let us recreate the graph with more features.

```{r recreate-graph}
madrid_sf_new <- opq(bbox = small_bbox) %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary", 
                            "tertiary", "residential", "unclassified",
                            "motorway_link", "trunk_link")) %>%
  osmdata_sf()

madrid_graph_components_new <- graph_components$new(madrid_sf_new, perform_merges = TRUE,tolerance = list(vertex_vertex = 0.001, vertex_edge = 0.01))

madrid_graph_new <- madrid_graph_components_new$get_largest()
```

## Examining the new graph along with the data

```{r examine-new-graph}
# Plot the new graph with the sensor locations
p <- madrid_graph_new$plot(vertex_size = 0, edge_width = 0.5)
p + geom_sf(data = radar_locations_sf, color = "red", size = 1)
```

We can now add the data to the graph.

```{r add-data-to-graph}
madrid_graph_new$add_observations(madrid_traffic_avg_sf)
```



## Exercise 2: Checking Coordinate Reference Systems

A crucial step in spatial analysis is ensuring that all data uses the same coordinate reference system (CRS):

```{r check-crs}
# Check the CRS of the radar locations
st_crs(radar_locations_sf)

# Check the CRS of the graph
print(madrid_graph)
```

## Examining Traffic Data

Now let's examine the pre-computed traffic measurements:

```{r examine-traffic}
# Basic statistics of the traffic data
summary(madrid_traffic_avg_sf)

# Histogram of traffic intensity
hist(madrid_traffic_avg_sf$intensity, 
     main = "Distribution of Traffic Intensity", 
     xlab = "Intensity (vehicles per 15min)",
     col = "lightblue")
```

# Visualizing Traffic Data on the Graph

Let's check if the traffic data is already loaded into the graph:

```{r check-graph-data}
  madrid_graph_new$add_observations(madrid_traffic_avg_sf)
  madrid_graph_new$add_observations(
    madrid_graph_new$mutate(log_intensity = log(pmax(1, intensity))), 
    clear_obs = TRUE
  )
  
  # Plot the data
  madrid_graph_new$plot(data = "log_intensity", 
                   vertex_size = 0, 
                   data_size = 2,
                   edge_width = 0.5) +
    scale_color_viridis_c(name = "Log Intensity")
```

# Exploratory Data Analysis

Let's explore the relationships between different traffic metrics:

```{r eda}
# Get the data from the graph
traffic_data <- madrid_graph_new$get_data()
# Histogram of log intensity
hist(traffic_data$log_intensity, 
     main = "Histogram of Log Traffic Intensity", 
     xlab = "Log Intensity", 
     col = "lightblue", 
     breaks = 20)
```

# Spatial Modeling

Now let's fit a spatial model to the log intensity using a Whittle-Matérn random field with $\alpha = 1$, which corresponds to an exponential covariance model:

```{r fit-model}
# Fit the model
fit_alpha1 <- graph_lme(log_intensity ~ 1, 
                        graph = madrid_graph_new, 
                        BC = 0,
                        model = list(type = "WhittleMatern", alpha = 1))

# Look at the model summary
summary(fit_alpha1)
```

## Interpreting the Model Results

The estimated parameters of the Whittle-Matérn model provide insights into the spatial correlation of traffic intensity:

```{r interpret-model}
# Extract parameters from the model
summary(fit_alpha1)
```

- **Intercept**: The average log traffic intensity across the network
- **Range**: The effective distance (in kilometers) at which spatial correlation becomes negligible
- **Sigma**: The standard deviation of the spatial process, indicating the variability of traffic intensity
- **Nugget**: The measurement error or micro-scale variation

## Predicting on the Graph

We can use our fitted model to make predictions at unobserved locations on the graph:

```{r predictions}
# Build a mesh on the graph for prediction (if not already done)
if (length(madrid_graph_new$mesh$VtE) == 0) {
  madrid_graph_new$build_mesh(h = 0.01)
}

# Create a data frame with the prediction locations
# Extract mesh vertices as sf object
mesh_vertices_df <- data.frame(
  edge_number = madrid_graph_new$mesh$VtE[,1],
  distance_on_edge = madrid_graph_new$mesh$VtE[,2]
)

# Make predictions
pred <- predict(fit_alpha1, newdata = mesh_vertices_df,
                normalize = TRUE)

# Plot the predictions
madrid_graph_new$plot_function(X = pred$mean, 
                          vertex_size = 0, 
                          edge_width = 0.5) +
  scale_color_viridis_c(name = "Predicted\nLog Intensity")
```

## Fitting the model with `inlabru` interface

Let us create the SPDE model object:

```{r create-spde}
spde_model <- graph_spde(madrid_graph_new, alpha = 1)
```

Now, let us create the data object:

```{r create-data}
data_spde <- graph_data_spde(spde_model, loc_name = "loc")
```

Let us now fit the model:

```{r fit-model-inlabru, warning=FALSE}
f.s <- log_intensity ~ Intercept(1) + 
                    field(loc, model = spde_model)
fit_alpha1_inlabru <- bru(f.s, 
                    data = data_spde[["data"]])
```

Now, let us look at the model summary:

```{r interpret-model-inlabru}
summary(fit_alpha1_inlabru)
```

Now, let us see the estimated parameters:

```{r interpret-model-inlabru-2}
spde_result <- spde_metric_graph_result(fit_alpha1_inlabru, 
                    "field", spde_model)

summary(spde_result)
```


# Exercise 3: Further analysis

* Consider removing the zero values from the data.
* Try pruning the graph.
* Try fitting a model with smoothness parameter alpha = 2.
* Try fitting the model estimating the smoothness parameter by using `rSPDE` interface.
*  Try fitting the model with `INLA` interface.
*  Check the connected components of the graph and consider whether more features should be added to the graph, or if the tolerances should be changed.